Towards Practical First-Order Model Counting

### 2025-02-02 doc/paper/paper.tex
Section 4.1

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{CC}$}
\newcommand{\Cranetwo}{\textsc{Gantry}}
\newcommand{\Cranebfs}{\textsc{Gantry-BFS}}
\newcommand{\Cranegreedy}{\textsc{Gantry-Greedy}}

@@

@@

Our empirical evaluation sought to compare the runtime performance of
{\Cranetwo} with the current state of the art, namely \textsc{FastWFOMC} and
\textsc{ForcLift}. Our experiments involve two versions of \Cranetwo{}:
\Cranegreedy{} and \Cranebfs{}. Like its predecessor, \Cranetwo{} has two modes
for applying compilation rules to formulas: one that uses a greedy search
algorithm similar to \textsc{ForcLift} and another that combines greedy and
BFS\@.

@@

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Make sure it's clear what the `raw' instances are
    \item Don't emphasise that the benchmarks come from my previous work
    \item Integrate the example below into the text, possibly moving irrelevant
          details (such as the formulas in other logics) to supplementary
          material
  \end{itemize}
}

We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by Dilkas and Belle~\cite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' (hitherto
referred to as `friends') Markov logic
network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In \Ctwo{},
\FO{}, and \UFO{}, this problem can be formulated as
\[
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\]
or, equivalently, in conjunctive normal form as
\[
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\]
Finally, we include the bijection-counting problem previously utilised by Dilkas
and Belle~\cite{DBLP:conf/kr/DilkasB23}. Its formulation in \FO{} is described
in \cref{example:bijections}.

\begin{example}[Counting functions]\label{example:functions}
  To define predicate $P$ as a function from a domain $\Delta$ to itself, in
  \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)) \land (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z).
    \end{gathered}
  \end{equation}
  The first clause asserts that each $x$ must have at least one corresponding
  $y$, while the second statement adds the condition that if $x$ is mapped to
  both $y$ and $z$, then $y$ must equal $z$. It is important to note that
  \cref{eq:fo} is written with two domains instead of just one. However, we can
  still determine the correct number of functions by assuming that the sizes of
  $\Gamma$ and $\Delta$ are equal. This formulation, as observed by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}, can prove beneficial in enabling FOKC
  algorithms to find efficient solutions.
\end{example}

@@

\todo[inline]{For each curve, check which polynomial or exponential model best
  fits the data to expand this section a bit (without plotting them, but
  describing the method).}

The three benchmark families cover a wide range of possibilities. The
`friends' benchmark stands out as it uses multiple predicates and can be
expressed in \FO{} using just two variables without cardinality constraints or
counting quantifiers. The `functions' benchmark, on the other hand, can still be
handled by all the algorithms, but it requires cardinality constraints, counting
quantifiers, or more than two variables. Lastly, the `bijections' benchmark is
an example of a formula that \textsc{FastWFOMC} can handle but \textsc{ForcLift}
cannot.

For evaluation purposes, we ran each algorithm on each benchmark using domains
of sizes $2^{1}, 2^{2}, 2^{3}$, and so on, until an algorithm failed to handle a
domain size due to timeout, out of memory error, or out of precision errors.
While we separately measured compilation and inference time, we primarily focus
on total runtime, dominated by the latter.

\todo[inline,caption={}]{
  \begin{itemize}
     \item We are not aware of any formulas on which \Cranetwo{} scales worse
          compared to either \textsc{ForcLift} or \textsc{FastWFOMC}. The one
          advantage that \textsc{FastWFOMC} has over \Cranetwo{} is its support
          for counting quantifiers.
    \item Regarding programming languages and accuracy, we verified that the
          answers match for smaller domain sizes. Also, although written in
          different languages, both \Cranetwo{} and \textsc{FastWFOMC} use the
          GNU Multiple Precision Arithmetic Library.
  \end{itemize}
}

@@

\Cref{fig:plot} presents a summary of the experimental results. Only
\textsc{FastWFOMC} and \Cranebfs{} could handle the bijection-counting problem.
For this benchmark, the largest domain sizes these algorithms could accommodate
were \num{64} and \num{4096}, respectively. On the other two benchmarks,
\textsc{ForcLift} had the lowest runtime. However, since it can only handle
model counts smaller than $2^{31}$, it only scales up to domain sizes of
\num{16} and \num{128} for `friends' and `functions', respectively.
\textsc{FastWFOMC} outperformed \textsc{ForcLift} in the case of `friends', but
not `functions', as it could handle domains of size \num{1024} and \num{64},
respectively. Furthermore, both \Cranebfs{} and \Cranegreedy{} performed
similarly on both benchmarks. Similarly to the `bijections' benchmark,
\Cranetwo{} significantly outperformed the other two algorithms, scaling up to
domains of size \num{8192} and \num{67108864}, respectively.

@@

In this work, we have presented a scalable automated FOKC-based approach to
FOMC\@. Our algorithm involves completing the definitions of recursive functions
and subsequently translating all function definitions into C++ code. Empirical
results demonstrate that \Cranetwo{} can scale to larger domain sizes than
\textsc{FastWFOMC} while supporting a wider range of formulas than
\textsc{ForcLift}. The ability to efficiently handle large domain sizes is
particularly crucial in the weighted setting, as illustrated by the `friends'
example discussed in \cref{sec:experiments}, where the model captures complex
social networks with probabilistic relationships. Without this scalability, the
practical usefulness of these models would be limited.

@@

`Friends' in \Ctwo{} and \UFO{} is the same as in \FO{}.

For `bijections', the equivalent formula in \Ctwo{} is
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{.
  }P(x, y)).
\]
Similarly, in \UFO{} the same formula can be written as
\[
  (\forall x, y \in \Delta\text{.
  }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{.
  }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.

For `functions', in \Ctwo{} one would write $\forall x \in \Delta\text{.
}\exists^{=1} y \in \Delta\text{. }P(x, y)$. In \UFO{}, the same could be
written as
\begin{equation}\label[formula]{eq:functions1}
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)) \land (|P| = |\Delta|),
\end{equation}
where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models compared to
its counterpart in \Ctwo{}, the negative weight $w^{-}(S) = -1$ makes some of
the terms in \cref{eq:wfomc} cancel out.

>>>>

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{CC}$}
\newcommand{\Cranetwo}{\textsc{Gantry}}
\newcommand{\Cranebfs}{\textsc{Gantry-BFS}}
\newcommand{\Cranegreedy}{\textsc{Gantry-Greedy}}
\newcommand{\friends}{\emph{Friends \& Smokers}}
\newcommand{\functions}{\emph{Functions}}
\newcommand{\bijections}{\emph{Bijections}}

@@

\todo[inline]{Reformulate the theorem above into an upper bound (as I informally
  mention earlier in the text)}

@@

Our empirical evaluation sought to compare the runtime performance of
{\Cranetwo} with the current state of the art, namely \textsc{FastWFOMC} and
\textsc{ForcLift}. Our experiments involve two versions of \Cranetwo{}:
\Cranegreedy{} and \Cranebfs{}. Like its predecessor (see
\cref{sec:compilation}), \Cranetwo{} has two modes for applying compilation
rules to formulas: one that uses a greedy search algorithm similar to
\textsc{ForcLift} and another that combines greedy and BFS\@.

@@

We ran each algorithm on each benchmark using domains of sizes
$2^{1}, 2^{2}, 2^{3}$, and so on, until an algorithm failed to handle a domain
size due to timeout (of \SI{1}{\hour}), out of memory error, or out of precision
errors. While we separately measured compilation and inference time, we
primarily focus on total runtime, dominated by the latter.

@@

We compare these algorithms using three benchmarks from previous work. The first
benchmark is the bijection-counting problem from \cref{example:bijections}. The
next benchmark is a variant of the well-known \friends{} Markov logic
network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}, which can be
formulated as
\[
  (\forall x,y \in \Delta\text{.
  } S(x) \land F(x, y) \Rightarrow S(y)) \land (\forall x \in \Delta\text{.
  }S(x) \Rightarrow C(x)).
\]
In this formula, we have three predicates $S$, $F$, and $C$ that denote smoking,
friendship, and cancer, respectively. The first clause states that friends of
smokers are also smokers, and the second clause asserts that smoking causes
cancer. Common additions to this formula include making the friendship relation
symmetric and assigning probabilities to each clause. Finally, we include the
function-counting problem~\cite{DBLP:conf/kr/DilkasB23}
\[
  (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{.
  }P(x, y)) \land (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{.
  }P(x, y) \land P(x, z) \Rightarrow y = z)
\]
as our last benchmark. Here, predicate $P$ is defined as a function from
$\Gamma$ to $\Delta$. The first clause asserts that each $x$ must have at least
one corresponding $y$, while the second clause ensures that there is only one
such $y$.

\begin{remark*}
  We formulate \bijections{} and \functions{} benchmarks using two domains
  $\Gamma$ and $\Delta$ as such a formulation is known to help FOKC algorithms
  find efficient solutions~\cite{DBLP:conf/kr/DilkasB23}. To compare \Cranetwo{}
  and \textsc{ForcLift} with \textsc{FastWFOMC} that has no support for multiple
  domains, we set $|\Gamma| = |\Delta|$.
\end{remark*}

The three benchmarks cover a wide range of possibilities. The \friends{}
benchmark uses multiple predicates and can be expressed in \FO{} using two
variables without cardinality constraints or counting quantifiers. The
\functions{} benchmark, on the other hand, can still be handled by all the
algorithms, but it requires cardinality constraints, counting quantifiers, or
more than two variables, depending on the formulation and the capabilities of
the algorithm. Lastly, the \bijections{} benchmark is an example of a formula
that \textsc{FastWFOMC} can handle but \textsc{ForcLift} cannot.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add exact powers of two as vertical dotted lines (with both
          labels and lines in, say, red). Also try to expand this section a bit.
    \item We are not aware of any formulas on which \Cranetwo{} scales worse
          compared to either \textsc{ForcLift} or \textsc{FastWFOMC}. The one
          advantage that \textsc{FastWFOMC} has over \Cranetwo{} is its support
          for counting quantifiers.
    \item Regarding programming languages and accuracy, we verified that the
          answers match for smaller domain sizes. Also, although written in
          different languages, both \Cranetwo{} and \textsc{FastWFOMC} use the
          GNU Multiple Precision Arithmetic Library.
  \end{itemize}
}

@@

\Cref{fig:plot} presents a summary of the experimental results. Only
\textsc{FastWFOMC} and \Cranebfs{} could handle the bijection-counting problem.
For this benchmark, the largest domain sizes these algorithms could accommodate
were \num{64} and \num{4096}, respectively. On the other two benchmarks,
\textsc{ForcLift} had the lowest runtime. However, since it can only handle
model counts smaller than $2^{31}$, it only scales up to domain sizes of
\num{16} and \num{128} for \friends{} and \functions{}, respectively.
\textsc{FastWFOMC} outperformed \textsc{ForcLift} in the case of \friends{}, but
not \functions{}, as it could handle domains of size \num{1024} and \num{64},
respectively. Furthermore, both \Cranebfs{} and \Cranegreedy{} performed
similarly on both benchmarks. Similarly to the \bijections{} benchmark,
\Cranetwo{} significantly outperformed the other two algorithms, scaling up to
domains of size \num{8192} and \num{67108864}, respectively.

@@

In this work, we have presented a scalable automated FOKC-based approach to
FOMC\@. Our algorithm involves completing the definitions of recursive functions
and subsequently translating all function definitions into C++ code. Empirical
results demonstrate that \Cranetwo{} can scale to larger domain sizes than
\textsc{FastWFOMC} while supporting a wider range of formulas than
\textsc{ForcLift}. The ability to efficiently handle large domain sizes is
particularly crucial in the weighted setting, as illustrated by the \friends{}
example discussed in \cref{sec:experiments}, where the model captures complex
social networks with probabilistic relationships. Without this scalability, the
practical usefulness of these models would be limited.

@@

\friends{} in \Ctwo{} and \UFO{} is the same as in \FO{}.

For \bijections{}, the equivalent formula in \Ctwo{} is
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{.
  }P(x, y)).
\]
Similarly, in \UFO{} the same formula can be written as
\[
  (\forall x, y \in \Delta\text{.
  }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{.
  }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.

For \functions{}, in \Ctwo{} one would write $\forall x \in \Delta\text{.
}\exists^{=1} y \in \Delta\text{. }P(x, y)$. In \UFO{}, the same could be
written as
\begin{equation}\label[formula]{eq:functions1}
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)) \land (|P| = |\Delta|),
\end{equation}
where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models compared to
its counterpart in \Ctwo{}, the negative weight $w^{-}(S) = -1$ makes some of
the terms in \cref{eq:wfomc} cancel out.
